---
title: "Exploring Redistricting Changes: California 2022–2026"
output: html_document
---

```{r setup, include=FALSE}
# suppress warnings and messages for cleaner output
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Load and Inspect Data

Load the datasets representing two redistricting plans.

```{r load-data}
# Replace these paths with your local copies
ca2022 <- read.csv('/Users/cervas/Library/CloudStorage/GoogleDrive-jcervas@andrew.cmu.edu/My Drive/GitHub/createMaps/CA/CA-2022.csv')
ca2026 <- read.csv('/Users/cervas/Library/CloudStorage/GoogleDrive-jcervas@andrew.cmu.edu/My Drive/GitHub/createMaps/CA/CA-2026.csv')

# Quick overview of structure
head(ca2022)
head(ca2026)
```

---

## 2. Compare Democratic-Majority Districts

Count the number of districts where the Democratic share exceeds 50%.

```{r majority-counts}
old_counts <- colSums(1 * (ca2022[,-1] > .5))
new_counts <- colSums(1 * (ca2026[,-1] > .5))

old_counts
new_counts
```

---

## 3. Identify District-Level Changes

Highlight districts that shifted from below 50% to above 50%.

```{r comparison-map}
# exclude ID column
cols <- setdiff(names(ca2022), "ID")

# identify where vote share crossed threshold
compare_mat <- (ca2022[, cols] < 0.5) & (ca2026[, cols] > 0.5)

# create color-coded matrices
comparison_df <- as.data.frame(ifelse(compare_mat, "#1375B7", "#EEEEEE"))
comparison_df_stroke <- as.data.frame(ifelse(compare_mat, "#000000", "none"))

# add back ID column
comparison_df <- cbind(ID = ca2022$ID, comparison_df)
comparison_df_stroke <- cbind(ID = ca2022$ID, comparison_df_stroke)

# clean column names
names(comparison_df) <- gsub(".", "", names(comparison_df), fixed = TRUE)
names(comparison_df_stroke) <- gsub(".", "", names(comparison_df_stroke), fixed = TRUE)
names(comparison_df_stroke)[-1] <- paste0(names(comparison_df_stroke)[-1], "stroke")

# preview
head(comparison_df)
```

You can export the results if you want to link them to map visualizations.

```{r export-csv, eval=FALSE}
write.csv(comparison_df, '/path/to/district_changes.csv', row.names = FALSE)
write.csv(comparison_df_stroke, '/path/to/district_changes_stroke.csv', row.names = FALSE)
```

---

## 4. Visualize Changes by Election

Compare the number of Democratic-majority districts under the two plans.

```{r barplot}
res <- data.frame(
  Election = names(old_counts),
  OldPlan = old_counts,
  NewPlan = new_counts
)

# Compute net change
res$Change <- res$NewPlan - res$OldPlan

barplot(res$Change,
        names.arg = res$Election,
        col = ifelse(res$Change > 0, "blue", "red"),
        las = 2,
        ylim = c(min(res$Change)-1, max(res$Change)+1),
        ylab = "Change in Democratic-majority districts",
        axes = FALSE)
axis(2, at = seq(min(res$Change)-1, max(res$Change)+1, 1))
abline(h = 0, lty = 2)
box()
```

### Summary Table of District Flips

```{r summary-table}
# Create summary table of district flips
flip_summary <- data.frame(
  Election = c(
    "2018 Attorneys General", "2022 Attorneys General",
    "2018 Governor", "2022 Governor",
    "2022 Lt. Governor", "2016 President",
    "2020 President", "2024 President",
    "2022 US Senator", "2024 US Senator",
    "2022 Sec. Of State", "2022 Treasurer", "Average"
  ),
  Commission = c(44, 38, 44, 38, 39, 44, 45, 41, 42, 39, 39, 37, 40.83),
  Proposition50Plan = c(48, 46, 47, 46, 47, 47, 48, 47, 48, 47, 48, 46, 47.08),
  DistrictsFlipped = c(4, 8, 3, 8, 8, 3, 3, 6, 6, 8, 9, 9, 6.25)
)

# Add blue dot visual column
flip_summary$Visual <- sapply(
  flip_summary$DistrictsFlipped,
  function(n) paste(rep("🔵", round(n)), collapse = "")
)

# Display as a formatted table
knitr::kable(
  flip_summary,
  align = "lccc",
  col.names = c("Election", "Commission Plan", "Proposition 50 Plan", "Districts Flipped", ""),
  caption = "Comparison of Democratic-Majority Districts by Election"
)
```

---

## 5. Seat–Vote Relationship

Load the custom function and plot empirical seat–vote relationships.

```{r sv-relationship-function, include=FALSE}
source('https://raw.githubusercontent.com/jcervas/R-Functions/refs/heads/main/sv-hyp.R')
```

```{r sv-relationship}
# visualize baseline
# sv_plot_base(theme = "dark")

sv_plot_base(theme = "light",
              xlab = "% of Two-Party Vote",
              ylab = "% of Two-Party Seats",
              prop.line = TRUE,
              box = TRUE,
              xlim = c(0.45, 1),
              ylim = c(0.45, 1),
              main = "California")

# plot actual outcomes
points(x = colMeans(ca2022)[-1], y = colMeans(1 * (ca2022 > .5))[-1], col = "darkred", pch=19)
points(x = colMeans(ca2026)[-1], y = colMeans(1 * (ca2026 > .5))[-1], col = "black", pch=19)

# add fitted curves
sv_curve(v = colMeans(ca2022)[-1], s = colMeans(1 * (ca2022 > .5))[-1], col = "darkred")
sv_curve(v = colMeans(ca2026)[-1], s = colMeans(1 * (ca2026 > .5))[-1], col = "black")
```

---

## 6. Questions for Exploration

1. Which elections gained or lost Democratic-majority districts?
2. How does the seat–vote curve change between 2022 and 2026?
3. What does a blue cell in the comparison table represent?
4. What can we conclude about the 2026 map by comparing it to 2022?
